Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-13 write-ups/PLDI/paper.tex
Shortened abstract.

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current irregular algorithms only work on today's
GPUs by identifying hardware compute units and scheduling workgroups
exclusively to these compute units.  As well as being non-portable,
this is unacceptable because the GPU cannot be used in the meantime
for high-priority tasks with soft real-time constraints,
such as graphics rendering.

We propose \emph{cooperative kernels}, an extension to the traditional
GPU programming model geared towards writing blocking algorithms.
Cooperative kernels enable fair scheduling of workgroups executing a
blocking algorithm, as well as co-scheduling of other tasks on the
GPU, via a small set of language extensions that create a contract
between the programmer and the scheduler.  We describe our cooperative
kernels programming model and its semantics in detail, and our
implementation of a portable prototype of cooperative kernels on top
of OpenCL 2.0.  We evaluate the technique by porting a set of
irregular work stealing and graph algorithms to our programming model
and evaluate the performance and responsiveness of the approach across
four GPUs from two vendors.  Because our implementation exploits no
vendor-specific hardware, driver or compiler support, our encouraging
experimental results provide a lower-bound on the efficiency with which
cooperative kernels can be implemented in practice.

@@

\paragraph{Irregular algorithms and persistent kernels}
%
There has been a lot of work on accelerating blocking irregular
algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
programming style for long-running kernels.  These approaches rely on
the occupancy-bound execution model, and aim to take ownership of the
GPU by scheduling as many workgroups as the GPU has compute units,
relying on fair scheduling between these workgroups.  These approaches
render the GPU unavailable for other tasks, and are not future-proof
because (as discussed throughout the paper) the occupancy-bound
execution model is not guaranteed by GPU programming models and is
unlikely to be guaranteed in practice by future GPU platforms.

\ADComment{Now say that we could update these approaches to use cooperation.}

@@

Lots of work on irregular algorithms (assumes fairness), and persistent threads (assumes OB model, which assumes a degree of fairness).  Our contribution can enable these to be safe.

Work on multi-tasking to improve throughput, etc., but this does not relate to fairness and wouldn't help blocking algorithms.

Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.

@@

\ADComment{Kepler preemption support?  Mention in related work?}

@@

\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
  implement preemption, i.e.\ shows that it can be achieved.
  Simulator only preemption.  Not sure whether their approach would
  give fair scheduling.}

\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
  keep a GPU busy by flooding it with work.  Would absolutely not
  stand up to preemption.}

>>>>

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  This issue is currently circumvented by
exploiting scheduling quirks particular to today's GPUs in a manner
that does not allow the GPU to be shared with other workloads (such as
graphics rendering tasks).  We propose \emph{cooperative kernels}, an
extension to the traditional GPU programming model geared towards
writing blocking algorithms.  Workgroups of a cooperative kernel are
fairly scheduled, supporting blocking algorithms, and co-scheduling of
other tasks is supported via a small set of language extensions that
create a contract between the programmer and the scheduler.  We
describe the semantics of the cooperative kernels programming model
and discuss our portable prototype implementation.  We evaluate the
technique by porting a set of irregular work stealing and graph
algorithms to our programming model and assessing the performance and
responsiveness of the approach across several GPUs.  Because our
implementation exploits no vendor-specific hardware, driver or
compiler support, our encouraging experimental results provide a
lower-bound on the efficiency with which cooperative kernels can be
implemented in practice.

@@

\paragraph{Irregular algorithms and persistent kernels}
%
There has been a lot of work on accelerating blocking irregular
algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
programming style for long-running kernels.  These approaches rely on
the occupancy-bound execution model, and aim to take ownership of the
GPU by scheduling as many workgroups as the GPU has compute units,
relying on fair scheduling between these workgroups.  These approaches
render the GPU unavailable for other tasks, and are not future-proof
because (as discussed throughout the paper) the occupancy-bound
execution model is not guaranteed by GPU programming models and is
unlikely to be guaranteed in practice by future GPU platforms.
%
\ADComment{Now say that we could update these approaches to use cooperation.}

\paragraph{GPU multitasking}
%
\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
  keep a GPU busy by flooding it with work.  Would absolutely not
  stand up to preemption.}

\paragraph{Preemption support for GPUs}
%
\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
  implement preemption, i.e.\ shows that it can be achieved.
  Simulator only preemption.  Not sure whether their approach would
  give fair scheduling.}
%
Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
%
\ADComment{Kepler preemption support?  Mention in related work?}

@@

@@

@@
